Innovations Report: Procedural Fur System

### 2018-03-18 docs/paper.tex
more on docs

The most popular of these is XGen, originally developed by Walt Disney Animation Studios~\cite{xgen}, which has been included in Autodesk's Maya since 2014. XGen uses a 'Collection' and 'Description' paradigm; Descriptions store user controlled parameters which affect fur primitives (curves for example) on a select area of a mesh, and Collections act as a container for these Descriptions. There are two primary methods offered for distributing primitives on a mesh, Guide Curves can be used to manually control the placement and shape of primitives by placing 'Guides' on the mesh surface, whereas 'Groomable Splines' randomly distirbutes splines over the mesh surface which can be manipulated in real time using viewport manipulators that mimic the brushing and styling of fur. In both modes, 'modifiers' can be added to the modifier stack to further adjust the visual outcome, adding features such as clumping, noise, and coiling. Attributes within these modifiers can be further controlled using expressions or texture maps.

Another commercially available solution is Peregrine*Labs' Yeti~\cite{yeti}, which is also offered as a plugin within Autodesk's Maya. Yeti offers a more procedural approach, giving the user a node graph interface in which to construct the desired look from individual components. For example, the most simple usage requires the user to first create a 'scatter' node to distribute points on the mesh surface, and then combine it with a 'grow' node to extrude curves from these points. Whilst at first this may seem unintuitive when compared to XGen, it offers greater flexibility and avoids the pitfalls associated with a destructive workflow.

@@

@@

Once curve primitives have been generated, they need to be passed onto curve operators, which perform most of the work towards creating an appealing visual output. I decided to create an abstract curve operator node, much like I had with the distributor node, which contains a virtual function for performing the curve manipulation. This allows for multiple different curve operators to be integrated into the system with ease, however for the scope of this assignment I chose only to implement curve operators to provide bending, application of noise, and clumping functionality.

@@

For shading of the meshes I opted to implement a physically based shading model, as decribed in~\cite{learnPBR}, however some alterations were necessary as my method of rendering curves does not provide usable surface normal information. Instead, the normalized direction in which the curve is leading from one vertex to the next, was substituted as the surface normal. Then, all of the dot product terms in the shading calculations which use the surface normal, were inversed. The result roughly approximates the shading qualities of thin cylinders by assuming that all vectors that are perpendicular to the direction of the curve are valid surface normals. Whilst this technique provides visually pleasing results, it is inaccurate and does not account for properties such as self-shadowing, or light scattering.

@@

After I had implemented some of the features Alin requested, I decided that the system was in a satisfactory state for the purposes of this assignment. The following figure~\ref{figure:screenshots} shows some memes

\begin{figure}[htbp]
\centering
\subfloat[]{\includegraphics[width=0.45\linewidth]{images/rick}}
\hfill
\subfloat[]{\includegraphics[width=0.45\linewidth]{images/bunny}}
\caption{\label{figure:screenshots} literal memes}
\end{figure}

\begin{figure}[htbp]\centering
\includegraphics[width=1.0\linewidth]{images/test_0007}
\caption{\label{figure:screenshot1} An example of the visual output that the system is capable of producing.}
\end{figure}

Results Text.

@@

insert stats comparing CPU/GPU compute

@@

\begin{figure}[htbp]\centering
\begin{center}
\begin{tabular}{||c|c||}
\hline
Mesh Triangle Count & GPU Time (ms)\\
\hline
\hline
16 (Triangle Strip) & 18 - 20\\
\hline
4968 (Stanford Bunny) & 33 - 45\\
\hline
14067 (Utah Teapot) & 76 - 92\\
\hline
\end{tabular}
\caption{\label{figure:fullPerformance} Performance metrics showing the time taken to perform a full graph evaluation on a complex setup, as seen in Figure~\ref{figure:screenshot}, for meshes of varying triangle counts.}
\end{center}
\end{figure}

@@

@@

Unfortunately my system lacks the ability to export curves data, making it virtually unusable within a production pipeline in its current state.

@@

\begin{lstlisting}[language=GLSL, label={lst:buffers}, caption={Extract from the Distributor compute shader, showing how the Mesh and Curves buffers are accessed.}]
struct Face
{
	vec4 position[3];
	vec4 normal[3];
	vec4 uv[3];
};

>>>>

The most popular of these is XGen~\cite{xgen}, originally developed by Walt Disney Animation Studios, which has been included in Autodesk's Maya since 2014. XGen uses a 'Collection' and 'Description' paradigm; Descriptions store user controlled parameters which affect fur primitives (curves for example) on a select area of a mesh, and Collections act as a container for these Descriptions. There are two primary methods offered for distributing primitives on a mesh, Guide Curves can be used to manually control the placement and shape of primitives by placing 'Guides' on the mesh surface, whereas 'Groomable Splines' randomly distirbutes splines over the mesh surface which can be manipulated in real time using viewport manipulators that mimic the brushing and styling of fur. In both modes, 'modifiers' can be added to the modifier stack to further adjust the visual outcome, adding features such as clumping, noise, and coiling. Attributes within these modifiers can be further controlled using expressions or texture maps.

Another commercially available solution is Yeti~\cite{yeti}, which is also offered as a plugin within Autodesk's Maya. Yeti offers a more procedural approach, giving the user a node graph interface in which to construct the desired look from individual components. For example, the most simple usage requires the user to first create a 'scatter' node to distribute points on the mesh surface, and then combine it with a 'grow' node to extrude curves from these points. Whilst at first this may seem unintuitive when compared to XGen, it offers greater flexibility and avoids the pitfalls associated with a destructive workflow.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=0.49\linewidth]{images/triangleStripDistribution}
\hfill
\includegraphics[width=0.45\linewidth]{images/triangleMesh}
\caption{\label{figure:distribution} Example distribution generated by my naive random distributor, note how the distribution stays relatively uniform despite variance in mesh density.}
\end{figure}

@@

Once curve primitives have been generated, they need to be passed onto curve operators, which perform most of the work towards creating an appealing visual output. I decided to create an abstract curve operator node, much like I had with the distributor node, which contains a virtual function for performing the curve manipulation. This allows for multiple different curve operators to be integrated into the system with ease, however for the scope of this assignment I chose only to implement curve operators to provide bending and clumping functionality.

@@

For shading of the meshes I opted to implement a physically based shading model, as decribed in~\cite{learnPBR}, however some alterations were necessary as my method of rendering curves does not provide usable surface normal information. Instead the normalized direction in which the curve is leading, from one vertex to the next, was substituted as the surface normal. Then, all of the dot product terms in the shading calculations which use the surface normal, were inversed. The result roughly approximates the shading qualities of thin cylinders by assuming that all vectors that are perpendicular to the direction of the curve are valid surface normals. Whilst this technique provides visually pleasing results, it is inaccurate and does not account for properties such as self-shadowing, or light scattering.

@@

After I had implemented some of the features Alin requested, I decided that the system was in a satisfactory state for the purposes of this assignment. The following Figures~\ref{figure:screenshots} and ~\ref{figure:screenshot_gui} show examples of the kind of visual output my system is capable of producing, and the accompanying node graph setup used to create them.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.45\linewidth]{images/rick}
\hfill
\includegraphics[width=0.45\linewidth]{images/bunny}
\caption{\label{figure:screenshots} Examples of visual output that my system is capable of producing.}
\end{figure}

\begin{figure}[htbp]\centering
\includegraphics[width=1.0\linewidth]{images/nodes}
\caption{\label{figure:screenshot_gui} The accompanying node graph setup used to create the images in Figure~\ref{figure:screenshots}.}
\end{figure}

@@

Figures~\ref{figure:distributionPerformance} and~\ref{figure:fullPerformance} show that my system is performant enough to provide real-time feedback, even when dense meshes and high primitive counts are used. Whilst Figure~\ref{figure:distributionPerformance} shows performance gains by utilisizing the GPU over the CPU, note that my CPU implementation is single-threaded, and further performance gains could be made by utilizing multithreading.

It is worth noting that my system is quite inefficient in regards to GPU memory consumption; The node graph setup in Figure~\ref{figure:screenshot_gui} uses 322 MiB of VRAM. This is a result of storing the state of each \texttt{GL\_SHADER\_STORAGE\_BUFFER} in GPU memory, for each of the curve operators used. Whilst this avoids the need for a full graph evaluation each time a parameter is changed, it effectively doubles VRAM usage each time a curve operator is added.

@@

% \begin{figure}[htbp]\centering
% \begin{center}
% \begin{tabular}{||c|c||}
% \hline
% Mesh Triangle Count & GPU Time (ms)\\
% \hline
% \hline
% 16 (Triangle Strip) & 18 - 20\\
% \hline
% 4968 (Stanford Bunny) & 33 - 45\\
% \hline
% 14067 (Utah Teapot) & 76 - 92\\
% \hline
% \end{tabular}
% \caption{\label{figure:fullPerformance} Performance metrics showing the time taken to perform a full graph evaluation on a complex setup, as seen in Figure~\ref{figure:screenshot_gui}, for meshes of varying triangle counts.}
% \end{center}
% \end{figure}

\begin{figure}[htbp]\centering
\begin{center}
\begin{tabular}{||c|c||}
\hline
Mesh Triangle Count & Evaluation Time (ms)\\
\hline
\hline
16 (Triangle Strip) & 69 - 84\\
\hline
4968 (Stanford Bunny) & 81 - 112\\
\hline
14067 (Utah Teapot) & 138 - 228\\
\hline
\end{tabular}
\caption{\label{figure:fullPerformance} Performance metrics showing the time taken to perform a full graph evaluation on a complex setup, as seen in Figure~\ref{figure:screenshot_gui}, for meshes of varying triangle counts. Note that over 500,000 primitives are being processed in this setup.}
\end{center}
\end{figure}

@@

Overall I am pleased with the visual quality and interactivity of my system, but I believe further work is necessary in order to maximise it's performance and usability.

% Currently my system performs all of its computation using OpenGL compute shaders, though better results may be attainable using other APIs such as CUDA or OpenCL.

@@

Unfortunately my system lacks the ability to export curves data, making it virtually unusable within a production pipeline in its current state. A viable file format for this could be \textit{Alembic}~\cite{alembic}, as it natively supports the storage of parametric curves. Integrating alembic support into my system could also allow for animated mesh geometry to be used.

Another desirable addition to my system would be viewport manipulators, similar to the 'Groomable Splines' feature in XGen. This could allow me to implement techniques such as those described in Skippy~\cite{skippy}, which would provide my system with a unique user experience compared to commercially available systems.

Whilst not necessary for usage in a broader pipeline, implementing a physically plausible hair shading model such as those described in \cite{Marschner}, would greatly improve the visual accuracy of my viewport renderings.

\newpage

@@

\begin{lstlisting}[language=GLSL, label={lst:buffers}, caption={Extract from the Distributor compute shader, showing how the Mesh and Curves buffers are accessed.}]
struct Face
{
	vec4 position[3];
	vec4 normal[3];
	vec4 uv[3];{}
};
